Interactive Filter and Display of Hillary Clinton's Emails: \\A Cautionary Tale of Metadata

### 2017-03-25 VAST_submission/vgtc_journal_latex/paper.tex
Paper update

The 2016 US Presidential election was one of the most contentious in history.  The existence and possible content of Hillary Clinton's private email server played an significant role in the campaign until the end. Despite their significant nature, very few of the individuals commenting on the significance of this server had spent any significant time viewing this content. This is no fault of theirs, the officially released data has previously only been stored in databases on the United States' State Department Website (see https://foia.state.gov/Learn/New.aspx) and Wikileaks (see https://wikileaks.org/clinton-emails/) in individually searchable form. Furthermore, the email data is stored in the inconvenient format of individual PDF documents, and in the case of Wikileaks a slightly more convenient but still cumbersome HTML representation based on the official PDF content. This granularity and separation of individual emails on different webpages prevents a great deal interesting analysis of the data, including any aggregate analyses of patterns in content or metadata. As such, the emails are incredibly misunderstood documents, and current searches for patterns and points of interest have relied upon searching for terms of interest and sifting through the results of the search. Such a method is not only tedious but problematic from a statistical point of view, as approaching any data set with a specific hypothesis before exploring its structure and patterns in abstract leads to biased conclusions and poor analysis. In the interest of simplifying the general exploration of this important data set, the Wikileaks HTML email versions were extracted and a series of visualizations were generated around the metadata included in these emails. An interactive web application in \texttt{R} \texttt{shiny} was constructed utilizing these visualizations and the extracted data. Incorporated in this app was a short article explaining the use of the app, providing links to background information, and outlining some of the findings of the more interesting trends observed. This framework provides any individual with interest the ability to generate hypotheses by exploring the data without necessarily searching for any particular conclusion. The methods used to generate this

@@

Learn:
\begin{itemize}
\item inner circle over time (state or not)
\item spike in email around Libyan revolution
\item gap in the email
\item daily email patterns, server behaviour
\end{itemize}

>>>>

The 2016 US Presidential election was one of the most contentious in history.  The existence and possible content of Hillary Clinton's private email server played an significant role in the campaign until the end. Despite their significant nature, very few of the individuals commenting on the significance of this server had spent any significant time viewing this content. This is no fault of theirs, the officially released data has previously only been stored in databases on the United States' State Department Website (see https://foia.state.gov/Learn/New.aspx) and Wikileaks (see https://wikileaks.org/clinton-emails/) in individually searchable form. Furthermore, the email data is stored in the inconvenient format of individual PDF documents, and in the case of Wikileaks a slightly more convenient but still cumbersome HTML representation based on the official PDF content. This granularity and separation of individual emails on different webpages prevents a great deal interesting analysis of the data, including any aggregate analyses of patterns in content or metadata. As such, the emails are incredibly misunderstood documents, and current searches for patterns and points of interest have relied upon searching for terms of interest and sifting through the results of the search. Such a method is not only tedious but problematic from a statistical point of view, as approaching any data set with a specific hypothesis before exploring its structure and patterns in abstract leads to biased conclusions and poor analysis. In the interest of simplifying the general exploration of this important data set, the Wikileaks HTML email versions were extracted and a series of visualizations were generated around the metadata included in these emails. An interactive web application in \texttt{R} \texttt{shiny} was constructed utilizing these visualizations and the extracted data. Incorporated in this app was a short article explaining the use of the app, providing links to background information, and outlining some of the findings of the more interesting trends observed. This framework provides any individual with interest the ability to generate hypotheses by exploring the data without necessarily searching for any particular conclusion.

In generating and exploring this data, a considerable amount of respect is gained for the power of the often overlooked metadata of emails, especially with public figures. Simply having access to the network of communications, simple indicators of identifiable features of the individuals communicating, and times sent is enough to generate interesting questions which simple Google searches of important events and days can shed a great deal of light on. Leveraging data beyond this metadata, including the Freedom of Information Act (FOIA) exemption codes (see https://vault.fbi.gov/explanation-of-exemptions) used to redact the emails, only bolsters this simple metadata driven investigation. Metadata is often viewed as less significant or somehow less intrusive data, but the exploration performed in this paper demonstrated, at the very least to the authors, what a powerful tool metadata is for generating hypotheses about data. Given the lack of privacy present in the modern internet age and the amount of information most individuals make publicly available, there is a clear warning here about the utility of metadata for bad or for good. For a public official like Hillary Clinton, the data become even more interesting due to a number of contentious issues which arose during her tenure as the Secretary of State, independent of and as a result of her use of a private server.

\section{Data Extraction}

Before any of the analysis could begin, the emails had to be converted from the individual PDF and HTML form into a more easily used data structure. The choice was made to use the Wikileaks HTML emails as the source for data extraction. This choice was made primarily due to the relative ease of loading the raw HTML pages on the Wikileaks database due to their regular method of storage (the email with ID <id> is stored at https://wikileaks.org/clinton-emails/emailid/<id>) and ease of loading. These qualities made the HTML emails amenable to programmatic extraction, which was completed using a host of web-scraping and string parsing tools in \texttt{R}. In particular the packages \texttt{RCurl} and \texttt{XML} proved invaluable to extract the raw HTML pages for each email, and the packages \texttt{tm}, \texttt{stringr}, and \texttt{SnowballC} were indispensible to process and clean the raw HTML into a useable form. After downloading the raw HTML and extracting the data of interest, the resulting data was stored in csv files to provide the flexibility to load the data into any architecture or analysis tool desired.

\section{Shiny App}

@@

Learn:
\begin{itemize}
\item inner circle over time (state or not)
\item spike in email around Libyan revolution
\item gap in the email
\item daily email patterns, server behaviour (daylight savings time)
\end{itemize}
